Are populations like a circuit? The relationship between isolation by distance and isolation by resistance

### 2018-08-23 isolation_by_coalescence.tex
minor edits by erik

For data, we are given genotypes of individuals and their geographic coordinates;
then, we divide these individuals into groups,
and compute the genetic distance between group $i$ and group $j$
as the mean divergence between a pair of genomes, chosen one from each group.
We denote the resulting genetic distance $D_{ij}$;
since divergence is symmetric, $D_{ij} = D_{ji}$.
We compute $D_{ii}$ by sampling without replacement from group $i$,
so that generally $D_{ii} > 0$: the diagonal elements measure local diversity.
As discussed below, we will find commute times so that the mean commute time between
$i$ and $j$ matches $D_{ij}$ -- but the commute time of a location to itself is zero,
so we must account for local diversity.
We do this in the same way as \citet{petkova2016visualizing},
by introducing another set of model parameters, $q$, 
that fill the same role as the coalescence rates, $\gamma$.
Then, we use commute time to approximate genetic distance as
\begin{align} \label{eq:commute_approx}
	D_{ij} \approx \comdist{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
\end{align}
The factor of four appears to make commute time more similar to coalescence time:
the commute time covers the path from $i$ to $j$ twice
(once in both directions),
while the coalescence time covers the distance only once,
and in half the time, because two particles are moving.
Furthermore, coalescence time includes within-location diversity,
because particles do not coalesce instantly --
commute time is most analogous to the first meeting time.
For a conceptual picture of the differences between the coalescence time and commute time,
see Figure \ref{fig:concept_coalcom}.

@@

Now, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the jump rate from $x$ to $y$.
This means that if the Markov chain is at location $X_t$ at time $t$,
then for $x \neq y$, the probability the chain jumps from $x$ to $y$ is proportional to $G_{xy}$,
i.e., $\PP{X_{t+\epsilon} = y \given X_t = y} = \epsilon G_{xy} + O(\epsilon^2)$,
and $G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain,
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
By conditioning on $X_\epsilon$,
one can use the Markov property to show that
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z,
\end{align}
i.e., if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$.

@@

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra
(and is related to the Sylvester equation \citep{sylvester_eqn}).
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZH$ is antisymmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $Z$
so that $G + Z$ remains the generator of a valid Markov chain,
but this is not guaranteed in general.

@@

This suggests that although nonuniqueness of solutions may not be a problem in practice,
as long as geography is discretized into sufficiently many locations.
However, another problem appears at finer geographic resolutions:
although solutions may be in principle unique,
the problems become \emph{ill-conditioned}, 
in the sense that arbitrarily small variations in the input times
(even numerical instability)
produce large differences in the inferred rates.
Very roughly speaking, this occurs because as the spatial resolution increases,
the matrix $G$ converges to a second-order elliptic differential operator \citep{pde};
such operators are deformations of the Laplacian \citep{laplacian},
which is well-known to have eigenvalues that decay exponentially \citep{laplacian_eigenvals}.
Information about the landscape contained in higher-order modes
(e.g., finer resolution changes in hitting times)
is then almost obliterated by the small eigenvalues of $G$,
making the inverse problem possible in theory but not in practice.
For more discussion of related problems, see \citet{badtruth,shape_of_a_pop,inverse_prob}.

\paragraph{When are coalescence and commute times equal?}
If in a given biological situation the two times were equal --
i.e., if given $G$ and $\gamma$, there was a $q$ that made $\comdist = C$ --
then inference with either model would be equivalent.
We have already shown that this is not the case in general,
but it turns out that it \emph{is} true
under some restrictive assumptions not uncommonly found in population genetics.
We show in Appendix \ref{sec:com_eq_coal} that
if hitting times are symmetric ($H = H^T$),
then this does occur,
and so the two methods are equivalent for symmetric island models
(where the populations are arranged on a ring 
and migration rates only depend on the distance between them).
Hitting times are not symmetric for a square grid with uniform migration
(the mean time to hit the center from a corner is les than the reverse),
but the square grid is quite close to the torus, which does have symmetric hitting times.

@@

Mean coalescence times estimated from real data are subject to a number of sources of noise
(which we discuss in more detail later).
Here, we use
the exact solutions based on linear algebra developed above
to develop a Bayesian inference method that accounts for noise in the data
and estimates uncertainty in the resulting estimates.
To do this, we model genetic distances as equal to coalescence times plus noise:
$D_{ij} = C_{ij} + \epsilon_{ij}$,
where $\epsilon_{ij}$ are independent and Normally distributed
with mean $\mathbf{0}$ and variance $\sigma_\epsilon^2$,
except that $\epsilon_{ij} = \epsilon_{ji}$.
(We imagine that $C_{ij}$ is large compared to $\sigma_{\epsilon}$,
so that resulting values are still positive.)
Using the calculations above, 
given $G$ and $\gamma$ we can compute the corresponding coalescence times
-- call these $\mathcal{C}(G, \gamma)$ --
so our model is that $D$ has a Gaussian distribution 
with mean $\mathcal{C}(G, \gamma)$.
The maximum likelihood estimate of $G$ could be found directly;
however, this in practice will quite likely contain negative movement rates,
and does not allow us to impose constraints
(such as only allowing a subset of movement rates to be nonzero).

Therefore, we use Bayesian inference,
placing independent exponential priors on the parameters --
on those movement rates $G_{ij}$ that are not constrained to be zero,
and on the coalescence rates, $\gamma$.
Concretely, suppose that the spatial arrangement of populations is given by a graph,
and only movement rates corresponding to edges in the graph are allowed to be nonzero.
Write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph,
that there are $m_G$ edges in the graph,
and that each of the $n$ populations has its own coalescence rate.
Then, the resulting log-posterior density is
\begin{align} \label{eq:post}
\mathcal{L}(D, G, \gamma) 
    &=
    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
	- \sum_{i \sim j} \lambda_G G_{ij}  \\
    &\qquad
    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
        \right)^2 
\end{align}
Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
respectively.

@@

It is not required in this approach to have samples from every spatial location,
allowing inference on incompletely sampled spatial discretizations.
To carry out inference in this situation,
we simply use Equation \ref{eq:post},
but only summing over observed entries of $D_{ij}$.

\paragraph{MCMC methods}
To sample from the posterior distribution of equation \eqref{eq:post}, 
we used R \citep{R} 
to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
Starting locations were chosen by sampling from the prior distribution.
Before the usual ``burn-in'' phase before sampling for inference begins,
we included a period of ``pre-burn-in'' which used the same MCMC procedure
with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
portion of parameter space.
Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
and $3 \time 10^6$ iterations of burn-in,
followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
although for small graphs fewer iterations were necessary.
We used standard methods to assess convergence and mixing.

@@

\paragraph{Forwards-time simulations}
To produce realistic data, we implemented forwards-time simulations using SLiM v3.0 \citep{haller2017},
with individuals living across continuous, two-dimensional space (sometimes with barriers),
from which we recorded genomic data.
We performed three types of simulation:
uniform migration on a continuous square landscape, 
uniform migration on a rectangular landscape with barriers to migration, 
and biased migration on a square landscape.
Each simulation had $10^4$ diploid, hermaphroditic individuals with nonoverlapping generations.
Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
truncated at $3 \sigma_d$.
Offspring are dispersed away from the mother using the same kernel,
conditioned to lie within the habitable area.
In each simulation, $\sigma_d$ was chosen 
so that there were an average of 0.25 individuals per $\sigma_d^2$ units of area.
In order to prevent excessive population clumpiness \citep{felsensten1975pain}, 
we also model local competition. 
To do this, we use the same truncated Gaussian kernel to define an ``interaction strength''
between each pair of nearby individuals;
then, the fitness of each individual is set to $\exp(-x/40)$, 
where $x$ is the sum of their interaction strengths with other individuals.
This reduces fitness for individuals in denser areas.
Individuals are chosen to reproduce proportional to their fitness.
Clumpiness is similar to what is seen in 
Figure \ref{fig:ind_locs_5x3b_1}

>>>>

For data, we are given genotypes of individuals and their geographic coordinates;
then, we divide these individuals into groups,
and compute the genetic distance between group $i$ and group $j$
as the mean divergence between a pair of genomes, chosen one from each group.
We denote the resulting genetic distance $D_{ij}$;
since divergence is symmetric, $D_{ij} = D_{ji}$.
We compute $D_{ii}$ by sampling without replacement from group $i$,
so that generally $D_{ii} > 0$: the diagonal elements measure local diversity.
As discussed below, we will find commute times so that the mean commute time between
$i$ and $j$ matches $D_{ij}$ -- but the commute time of a location to itself is zero,
so we must account for local diversity.
We do this in the same way as \citet{petkova2016visualizing},
by introducing another set of model parameters, $q$, 
that fill the same role as the coalescence rates, $\gamma$.
Then, we use commute time to approximate genetic distance as
\begin{align} \label{eq:commute_approx}
	D_{ij} \approx \comdist_{ij} = R_{ij}/4 + q_{i}/2 + q_{j}/2 .
\end{align}
The factor of four appears to make commute time more similar to coalescence time:
the commute time covers the path from $i$ to $j$ twice
(once in both directions),
while the coalescence time covers the distance only once,
and in half the time, because two particles are moving.
Furthermore, coalescence time includes within-location diversity
because particles do not coalesce instantly --
therefore commute time is most analogous to the first meeting time.
For a conceptual picture of the differences between the coalescence time and commute time,
see Figure \ref{fig:concept_coalcom}.

@@

Now, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the jump rate from $x$ to $y$.
This means that if the Markov chain is at location $X_t$ at time $t$,
then for $x \neq y$, the probability the chain jumps from $x$ to $y$ is proportional to $G_{xy}$,
i.e., $\PP{X_{t+\epsilon} = y \given X_t = y} = \epsilon G_{xy} + O(\epsilon^2)$,
and $G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain --
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
By conditioning on $X_\epsilon$,
one can use the Markov property to show that
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z,
\end{align}
i.e., if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$.

@@

Equation \eqref{eqn:C_matrix} is linear in $G$ and $\gamma$,
and so after rearrangement can be solved with standard linear algebra
(and is related to the Sylvester equation \citep{sylvester_eqn}).
However, the solution is again not necessarily unique:
suppose that we have a matrix $Z$ such that $ZC$ is antisymmetric.
Then $(G + Z) C + C (G + Z)^T = GC + CG^T$,
and so a Markov chain with generator matrix $G + Z$ has the same coalescence times
as the original chain.
As before, if all entries of $G$ are nonzero, 
it is always possible to find sufficiently small $Z$
so that $G + Z$ remains the generator of a valid Markov chain,
but this is not guaranteed in general.

@@

This suggests that although nonuniqueness of solutions may not be a problem in practice,
as long as geography is discretized into sufficiently many locations.
However, another problem appears at finer geographic resolutions:
although solutions may be in principle unique,
the problems become \emph{ill-conditioned} 
in the sense that arbitrarily small variations in the input times
(even numerical instability)
produce large differences in the inferred rates.
Very roughly speaking, this occurs because as the spatial resolution increases,
the matrix $G$ converges to a second-order elliptic differential operator \citep{pde};
such operators are deformations of the Laplacian \citep{laplacian},
which is well-known to have eigenvalues that decay exponentially \citep{laplacian_eigenvals}.
Information about the landscape contained in higher-order modes
(e.g., finer resolution changes in hitting times)
is then almost obliterated by the small eigenvalues of $G$,
making the inverse problem possible in theory but not in practice.
For more discussion of related problems, see \citet{badtruth,shape_of_a_pop,inverse_prob}.

\paragraph{When are coalescence and commute times equal?}
If in a given biological situation the two times were equal --
i.e., if given $G$ and $\gamma$, there was a $q$ that made $\comdist = C$ --
then inference with either model would be equivalent.
We have already shown that this is not the case in general,
but it turns out that it \emph{is} true
under some restrictive assumptions not uncommonly found in population genetics.
We show in Appendix \ref{sec:com_eq_coal} that
if hitting times are symmetric ($H = H^T$),
then this does occur,
and so the two methods are equivalent for symmetric island models
(where the populations are arranged on a ring 
and migration rates only depend on the distance between them).
Hitting times are not symmetric for a square grid with uniform migration
(the mean time to hit the center from a corner is less than the reverse),
but the square grid is quite close to the torus, which does have symmetric hitting times.

@@

Mean coalescence times estimated from real data are subject to a number of sources of noise
(which we discuss in more detail later).
Here, we use
the exact solutions based on linear algebra developed above
to develop a Bayesian inference method that accounts for noise in the data
and estimates uncertainty in the resulting estimates.
To do this, we model genetic distances as equal to coalescence times plus noise:
$D_{ij} = C_{ij} + \epsilon_{ij}$,
where $\epsilon_{ij}$ are independent and Normally distributed
with mean $\mathbf{0}$ and variance $\sigma_\epsilon^2$,
except that $\epsilon_{ij} = \epsilon_{ji}$.
(We imagine that $C_{ij}$ is large compared to $\sigma_{\epsilon}$,
so that resulting values are still positive.)
Using the calculations above, 
given $G$ and $\gamma$ we can compute the corresponding coalescence times
-- which we call $\mathcal{C}(G, \gamma)$ --
so our model is that $D$ has a Gaussian distribution 
with mean $\mathcal{C}(G, \gamma)$.
The maximum likelihood estimate of $G$ could be found directly;
however, this in practice will quite likely contain negative movement rates,
and does not allow us to impose constraints
(such as only allowing a subset of movement rates to be nonzero).

Therefore, we use Bayesian inference,
placing independent exponential priors on the parameters --
the movement rates, $G_{ij}$, that are not constrained to be zero
and the coalescence rates, $\gamma$.
Concretely, suppose that the spatial arrangement of populations is given by a graph,
and only movement rates corresponding to edges in the graph are allowed to be nonzero.
Write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph,
that there are $m_G$ edges in the graph,
and that each of the $n$ populations has its own coalescence rate.
Then, the resulting log-posterior density is
\begin{align} \label{eq:post}
\mathcal{L}(D, G, \gamma) 
    &=
    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
	- \sum_{i \sim j} \lambda_G G_{ij}  \\
    &\qquad
    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
        \right)^2 
\end{align}
Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
respectively.

@@

It is not required in this approach to have samples from every spatial location,
allowing inference on incompletely sampled spatial discretizations.
To carry out inference in this situation,
we simply use Equation \ref{eq:post},
but only sum over observed entries of $D_{ij}$.

\paragraph{MCMC methods}
To sample from the posterior distribution of equation \eqref{eq:post}, 
we used R \citep{R} 
to implement a standard Metropolis-Hastings algorithm with Gaussian proposals \citep{brooks2011handbook}.
Starting locations were chosen by sampling from the prior distribution.
Before the usual ``burn-in'' phase before sampling for inference begins,
we included a period of ``pre-burn-in'' which used the same MCMC procedure
with a larger $\sigma_\epsilon$, to allow the chain to more quickly converge on the high-posterior
portion of parameter space.
Typically, we ran MCMC for $10^6$ iterations of pre-burn-in 
and $3 \times 10^6$ iterations of burn-in,
followed by $4 \times 10^6$ iterations which were used to estimate posterior distributions,
although for small graphs fewer iterations were necessary.
We used standard methods to assess convergence and mixing.

@@

\paragraph{Forwards-time simulations}
To produce realistic data, we implemented forwards-time simulations using SLiM v3.0 \citep{haller2017},
with individuals living across continuous, two-dimensional space (sometimes with barriers),
from which we recorded genomic data.
We performed three types of simulation:
uniform migration on a continuous square landscape, 
uniform migration on a rectangular landscape with barriers to migration, 
and biased migration on a square landscape.
Each simulation had $10^4$ diploid, hermaphroditic individuals with nonoverlapping generations.
Mates were selected from nearby individuals using a Gaussian kernel with standard deviation $\sigma_d$,
truncated at $3 \sigma_d$.
Offspring are dispersed away from the mother using the same kernel,
conditioned to lie within the habitable area.
In each simulation, $\sigma_d$ was chosen 
so that there were an average of 0.25 individuals per $\sigma_d^2$ units of area.
In order to prevent excessive population clumpiness \citep{felsensten1975pain}, 
we also model local competition. 
To do this, we use the same truncated Gaussian kernel to define an ``interaction strength''
between each pair of nearby individuals;
then, the fitness of each individual is set to $\exp(-x/40)$, 
where $x$ is the sum of their interaction strengths with other individuals.
This reduces fitness for individuals in denser areas.
Individuals are chosen to reproduce proportional to their fitness.
Clumpiness is similar to what is seen in 
Figure \ref{fig:ind_locs_5x3b_1}.
